---
title: "Climatic classification of drought typologies"
author: "Robert Monjo, Dominic Royé, Javier Martin-Vide"
date: "19 juny de 2019"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


## Abstract

Drought duration strongly depends on the definition thereof. In meteorology, dryness is habitually measured by means of fixed thresholds (e.g. 0.1 or 1 mm usually define dry spells) or climatic mean values (as is the case of the Standardised Precipitation Index), but this also depends on the aggregation time interval considered. However, robust measurements of drought duration are required for analysing the statistical significance of possible changes. Herein we have climatically classified the drought duration around the world according to their similarity to the voids of the Cantor set. Dryness time structure can be concisely measured by the n-index (from the regular/irregular alternation of dry/wet spells), which is closely related to the Gini index and to a Cantor-based exponent. This enables the world’s climates to be classified into six large types based upon a new measure of drought duration. We performed the dry-spell analysis using the full global gridded daily Multi-Source Weighted-Ensemble Precipitation (MSWEP) dataset. The MSWEP combines gauge-, satellite-, and reanalysis-based data to provide reliable precipitation estimates. The study period comprises the years 1979-2016 (total of 45165 days), and a spatial resolution of 0.5º, with a total of 259,197 grid points.


## Example of n-index applied to dry spells

### Loading functions
```{r}
source("n-index.r")
source("functions_spell.R")
```

### Load data example

The sample is three different points from the Multi-Source Weighted-Ensemble Precipitation (MSWEP) dataset on which is based the study.

```{r}
load("test_points.RData")
str(test_points)

#coordinates of sample points
test_points$lonlat
```

```{r}
pr <- test_points$pr

apply(pr[, 2:4], 2, mean_dw_days)

apply(pr[, 2:4], 2, nindex_spell)

apply(pr[, 2:4], 2, mean_spells)

```


## Map of drought typologies

### Required packages

```{r}
library(tidyverse)
library(sf)
library(raster)
library(rnaturalearth)
```

### Preparing data

1. Import the raster and the legend classes.
2. Project the raster from WGS84 to Robinson.
3. Joining the id classes to the drought clasification. 
4. Mapping the final data with ggplot2

```{r}
#raster data
drought_raster <- raster("./data/drought_class.tif")

#projection 
crs_robin <- "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"

#project
drought_raster <- projectRaster(drought_raster, 
                                crs = crs_robin,
                                method = "ngb",
                                over = TRUE)

#raster to xyz
drought_df <- rasterToPoints(drought_raster) %>%
                      as.data.frame() %>% 
                         mutate(id = as.character(drought_class)) %>%
                           dplyr::select(-drought_class)

#legend
drought_legend <- read_csv("./data/legend_drought_class.csv") %>%
                        mutate(id = as.character(id))


#join xyz with legend classes 
drought_df <- left_join(drought_df, drought_legend,  by = "id") 

#drought class to factor 
lab <- unique(drought_df$class)

drought_df <- mutate(drought_df, class = factor(class, lab[c(2,6, 1,3, 4:5)]))

```

### Mapping

```{r}
#world coastline
world <- ne_coastline(scale = 10, returnclass = "sf") %>% 
              st_transform(crs = crs_robin)

#breaks and labels for long/lat
brk_x <- c(-180,-130,-80,-30,0,30,80,130,179)
brk_y <- c(-90,-65,-40,-15,0,15,40,65,90)

lab_x <- case_when(brk_x<0 ~ str_c(abs(brk_x),"ºW"),
                   brk_x==0 ~ "0º",
                   brk_x>0 ~ str_c(brk_x,"ºE"))

lab_y <- case_when(brk_y<0 ~ str_c(abs(brk_y),"ºS"),
                   brk_y==0 ~ "0º",
                   brk_y>0 ~ str_c(brk_y,"ºN"))


#colours for legend
col <- c("#a50f15","#de2d26","#ffeda0","#d9ef8b","#31a354","#006837")


#graticules 
grat_map <- st_graticule() %>% 
             st_transform(crs_robin) %>% 
               st_geometry 


#map with ggplot2

ggplot(drought_df)+
  geom_tile(aes(x, y, fill = class))+
  geom_sf(data = world, 
          size = 0.3, 
          colour = "black")+
  geom_sf(data = grat_map,
          linetype = "dashed",
          colour = "grey70",
          alpha = 0.3, size = 0.15)+
  scale_fill_manual(values = col)+
     coord_sf()+
      theme_void()+
       labs(fill = "")

```



## How to cite

Data: Monjo, R; Royé, D; Martin-Vide, J. (2019). Drought lacunarity around the world and its classification. doi: [10.5281/zenodo.3247041](10.5281/zenodo.3247041)

n-index: 1) Monjo, R. (2016): Measure of rainfall time structure using the dimensionless n-index, Clim. Res., 67, 71–86. 2) Monjo, R. and Martin-Vide, J. (2016): Daily precipitation concentration around the world according to several indices, Int. J. Climatol., 36, 3828–3838.

